---
title: "Activity Monitoring data analysis"
output:
  html_document: 
    keep_md: true
---

```{r globals, echo=FALSE,results='hide'}
library(rmarkdown)
library(knitr)
library(dplyr)
library(xtable)
library(ggplot2)
opts_chunk$set(echo = TRUE, results = "asis")
```

First of all, lets load the data and convert the columns to the required data types.

```{r Q1, cache=TRUE, results='hide'}
library(xtable)
library(dplyr)
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip")
actFile <- unzip("activity.zip")
actDf <- read.table(actFile, header = TRUE, sep = ",", na.strings = "NA", colClasses = c("numeric","character","numeric"))
actDf$date <- as.Date(actDf$date,"%Y-%m-%d")
```

Let us analyse more about the number of steps taken on each day.
The below histogram dsplays the data distribution of the total number of steps taken on each day

```{r Q2}
actT <- tbl_df(actDf)
actSummary <- actT %>%
  group_by(date) %>%
  summarise(dt_sum = sum(steps, na.rm = TRUE), dt_mean = mean(steps, na.rm = TRUE), dt_median = median(steps, na.rm = TRUE))
hist(actSummary$dt_sum, xlab = "Total steps each day", main = "Histogram of the total number of steps taken each day")

```

The below table displays the mean and median of the total number of steps taken on each day.

```{r Q3}
subActSum <- actSummary[,c(1,3,4)]
names(subActSum) <- c("Date", "Mean Steps", "Median Steps")
xt <- xtable(subActSum)
xt$Date <- as.character(xt$Date)
print(xt, type="html")
```

Let us now try to understand the average daily activity pattern
Below is a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.
```{r Q4.Q5}
actIntSummary <- actT %>%
  group_by(interval) %>%
  summarise(int_sum = sum(steps, na.rm = TRUE), int_mean = mean(steps, na.rm = TRUE), int_median = median(steps, na.rm = TRUE))
plot(actIntSummary$interval, actIntSummary$int_mean, type = "l", col = "red", xlab = "Time Interval", ylab = "Average number of steps")
actIntOrder <- head(arrange(actIntSummary, desc(int_sum)),1)
maxInt <- actIntOrder$interval
maxStep <- actIntOrder$int_mean
```
The `r maxInt` 5-minute interval has the maximum number of average steps (`r maxStep`)

Note that there are a number of days/intervals where there are missing values. 
```{r Q6}
cntNAs <- sum(is.na(actDf$steps))
```
There are `r cntNAs' records with missing step counts for a given day/interval. We will now try to fill all of the missing values in the dataset with the the mean for that 5-minute interval to build the final tidy dataset.

```{r}
actMerged <- merge(actDf,actIntSummary, by.x = "interval", by.y = "interval")

for(i in 1:nrow(actMerged)){
  if(is.na(actMerged$steps[i])) {
    actMerged$steps[i] <- actMerged$int_mean[i]
  }
}
actFinal <- actMerged[,c(1:3)]
```

Below is a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
actFinalT <- tbl_df(actFinal)
actFinalSumm <- actFinalT %>%
  group_by(date) %>%
  summarise(dt_sum = sum(steps, na.rm = TRUE), dt_mean = mean(steps, na.rm = TRUE), dt_median = median(steps, na.rm = TRUE))
hist(actFinalSumm$dt_sum, xlab = "Total steps each day", main = "Histogram of the total number of steps taken each day")

```

Let us now assess the impact of imputing the missing data.
Below is the summary statistics of the original data set which had missing values
```{r}
summary(actSummary$dt_sum)
```

Below is the summary statistics of the final data set after imputing the missing values.
```{r}
summary(actFinalSumm$dt_sum)
```

We conclude that there is not much significance difference after imputing the NAs.

Finally, let us analyse differences in activity patterns between weekdays and weekends

```{r}
actFinal$day_type <- weekdays(actFinal$date)
setDayType <- function(x){
  if(x == "Saturday" | x == "Sunday"){
    "Weekend"
  } else "Weekday"
}
actFinal$day_type <- sapply(actFinal$day_type,setDayType)
actFinal$day_type <- factor(actFinal$day_type, levels = c("Weekday","Weekend"))
g <- ggplot(actFinal, aes(interval, steps))
g + geom_line(color = "blue") + facet_grid(day_type~.) + xlab("Interval") + ylab("Number of steps")
```